\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping

### 2018-10-10 revision/response.tex

\newcounter{QuestionNum}[ReviewerNum]
\setcounter{QuestionNum}{0}

\newcounter{EditorNum}
\setcounter{EditorNum}{1}

\newcounter{EdQuestionNum}[EditorNum]
\setcounter{EdQuestionNum}{0}

\newcommand{\ReviewerQuestion}[2]{
  \vspace{5pt}\goodbreak
  \noindent\fbox{Reviewer:~#1} #2
  \normalfont\par
}
\newcommand{\EditorQuestion}[1]{
  \vspace{5pt}\goodbreak
  \noindent\fbox{Editor:} #1
  \normalfont\par
}
\newcommand{\Response}[1]{
  \goodbreak
  \textcolor{blue!50!black}{#1}
  \normalfont\par
}

\newcommand{\Line}[1]{\textbf{line #1}\xspace} % to highlight for our use, in case line numbers change significantly and we have to correct.
\newcommand{\Lines}[1]{\textbf{line #1}\xspace} %
\newcommand{\CortyRPaper}{Corty and Valdar (2018+) ``vqtl: An R package for mean-variance QTL mapping'' G3/2018/200159\xspace}
\newcommand{\CortyReanalysisPaper}{Corty \emph{et. al.} (2018+) ``Mean-variance QTL mapping identifies novel QTL for circadian activity and exploratory behavior in mice'' G3/2018/200194\xspace}
\newcommand{\CortyMethodsPaper}{Corty and Valdar (2018+) ``QTL mapping on a background of variance heterogeneity'' G3/2018/200195\xspace}

@@

\begin{document}
\reversemarginpar

\section*{Introduction}

This response refers to paper number \textbf{3} of the 3 co-submitted papers:
\begin{enumerate}
\item \CortyMethodsPaper
\item \CortyReanalysisPaper
\item \textbf{\CortyRPaper}
\end{enumerate}

@@

\section*{Editor Comments}

\subsection*{Multi-manuscript Issues (identical across responses to reviewers)}

\EditorQuestion{
  Now that you have all three manuscripts potentially going forward for publication in G3, they can be more tightly linked.
}
\Response{
  We have updated the manuscripts such that each one refers to the other two in the introduction and discussion. We have also edited the introductions so that they are less duplicative and each is more clearly focused on its own distinct contributions.
}

\EditorQuestion{
   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the co-submitted manuscript, \CortyReanalysisPaper.
  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
}

@@

% Ibáñez N, Varona L, Sorensen D, Noguera JL. A study of heterogeneity of environmental variance for slaughter weight in pigs. Animal. 2007;2:19-26.

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Reviewer 1 Comments}

\ReviewerQuestion{1.1}{
  The authors mention ``in almost all traits of interest in human health and disease''.
  The detection of QTL has also been widely used in plant and animal populations.
}
\Response{
  The reviewer is quite correct that QTL mapping has been productively applied in plant an animal populations in addition to human populations.
  Rather than add references from plant and animal genetics to the introduction, we recognized that the introduction started unnecessarily broadly and therefore removed the first sentence.
  To address the reviewer's concern, however, we have added references to the next paragraph in the introduction (where the heritability of phenotype variance is discussed) that rely heavily on studies of model organisms, livestock and crops.
}

\ReviewerQuestion{1.2}{
  The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
  I would like to see an example with a larger genotype data set.
}
\Response{
  The unrealistically-small genome we used to illustrate the standard workflow serves an important purpose --- to allow a reader to replicate the procedure and its ``findings'' on a typical desktop in a matter of minutes.
  Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of real data using package \texttt{vqtl}, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
  Additionally, in response to other reviewer comments, we have added performance benchmarks to this manuscript, which make use of the dataset from Kumar et al 2013, as described below.
}

\ReviewerQuestion{1.3}{
  The CPU time is indicated for the data set of the example.
  I would like to have the CPU time information in a more realistic data set.
  Perhaps authors can use a free data set from the web.
}
\Response{
  Thank you for this recommendation.
  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package --- we have added two sets of benchmarks.
}
\Response{
  First, we demonstrate the performance of the package according to the number of computer cores used and the number of permutations conducted.
  For this set of benchmarks, we used the intercross from Kumar et. al 2013, which we reanalyzed in \CortyReanalysisPaper.
  This intercross population has 244 mice and was genotyped at 93 markers.
  After applying the hidden Markov model in \texttt{R/qtl} with \texttt{step.size} of 2, the number of loci to be tested was 582.
  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  This addition can be found on page 4.
}
\Response{
  Second, we demonstrate the performance of the package according to the size of the mapping population and the number of genetic markers, holding constant the number of permutations at 1000 and the number of cores at 32.
  For this set of benchmarks, we used crosses simulated with \texttt{R/qtl} so we could control the population size and number of genetic markers.
  This addition can be found on page 4.
}

\ReviewerQuestion{1.4}{
  In the simulation, the authors should indicate the magnitude of the genetic variance explained by the QTL.
}
\Response{
  Thank you for this recommendation.
  We have added this information to the manuscript and it now appears on page 2.
}

\ReviewerQuestion{1.5}{
  I think the authors should also indicate if the software is capable in terms of computing power to analyze the data from SNP genotyping devices or even sequencing.
}

### 2018-10-10 Package_vqt_Manuscript.tex
Updated formatting of co-submitted papers and response to match that of BVH

We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL (\CortyMethodsPaper). 
In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL (\CortyReanalysisPaper).

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
% \WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
%
%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
%
%The R package \texttt{vqtl} implements mean-variance QTL mapping
which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
%We introduce the \texttt{vqtl} package as follows.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

@@

To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.

@@

For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
For a given number of cores, the relationship between time and the number of permutations is linear (\autoref{fig:benchmark_kumar}), the slope depending on the number of cores and ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.

@@

We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in an F2 intercross.
This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
In the case of genetic factors, they can be mapped, as illustrated in one companion article (\CortyReanalysisPaper).
In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as illustrated in another companion article (\CortyMethodsPaper).

@@

\bibliography{10_Aim1,wv}

>>>>

We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL \citep{CortyValdar2018bvh-biorxiv}.
In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL \citep{Corty2018biorxiv}.

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
% \WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
%
%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\citep{CortyValdar2018bvh-biorxiv}),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
%
%The R package \texttt{vqtl} implements mean-variance QTL mapping
which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
%We introduce the \texttt{vqtl} package as follows.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \citet{Corty2018biorxiv} to report how computation time varies with marker density and number of permutations.

@@

To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure \citep{CortyValdar2018bvh-biorxiv}.
Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.

@@

For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \citet{Corty2018biorxiv}, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
For a given number of cores, the relationship between time and the number of permutations is linear (\autoref{fig:benchmark_kumar}), the slope depending on the number of cores and ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.

@@

We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in an F2 intercross.
This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
In the case of genetic factors, they can be mapped, as illustrated in one companion article \citep{Corty2018biorxiv}.
In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as illustrated in another companion article (\citep{CortyValdar2018bvh-biorxiv}).

@@

\bibliography{vqtl}
